# Week 1 - Home

Open the data file LifeSat.sav. 

```{r, message=FALSE}
library(foreign)
data <- read.spss("LifeSat.sav", to.data.frame = TRUE)
```

### Question 1.a

Make a descriptives table for the variables: LifSat, educ, ChildSup, SpouSup, and age. 

What is the average age in the sample? And the range (youngest and oldest child)? 

*Hint: Use* `library(psych); describe(); []`

<details>
  <summary>Click for explanation</summary>
The package `psych` contains many functions for exploring data. Install and load the package, then use the `describe()` function to describe the data:

```{r}
library(psych)
describe(data[, c("LifSat", "educ", "ChildSup", "SpouSup", "age")])
```
<\details>


### Question 1.b

Perform a simple regression with LifSat as the dependent variable and educ as the independent variable.

<details>
  <summary>Click for explanation</summary>

```{r}
results <- lm(LifSat ~ educ, data)
summary(results)
```
<\details>

### Question 1.c.

Do the same with age as the independent variable. 

<details>
  <summary>Click for explanation</summary>

```{r}
results <- lm(LifSat ~ age, data)
summary(results)
```
<\details>


### Question 1.d. 

Again with ChildSup as the independent variable. 

<details>
  <summary>Click for explanation</summary>

```{r}
results <- lm(LifSat ~ ChildSup, data)
summary(results)
```
<\details>


### Question 1.e.

Perform a multiple regression with LifSat as the dependent variable and educ, age and ChildSup as the independent variables. 

<details>
  <summary>Click for explanation</summary>

```{r}
results <- lm(LifSat ~ educ + age + ChildSup, data)
summary(results)
```
<\details>


### Question 1.f. 

Compare the results under 1.e with those obtained under 1.b-1.d. What do you notice when you compare the regression parameter for each of the three predictors in the multiple regression with the corresponding regression parameters obtained in the simple regressions? 
