# Week 2 - Reading questions

```{r include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
config <- yaml::yaml.load_file("config.yml")
```

#### Reference {-}

[Preacher, Kristopher J. and MacCullum, Robert C. (2003) Repairing Tom Swift’s Electric Factor Analysis Machine, Understanding Statistics 2(1) 13-43.](http://kuscholarworks.ku.edu/dspace/bitstream/1808/1492/1/preacher_maccallum_2003.pdf)

#### Questions {-}

1. What is a latent variable? 
1. What is factor analysis and what can you investigate using this method? 
1. In the introduction, Preacher and Maccallum talk about a “little jiffy” method of doing factor analysis. Can you shortly list what this litlle jiffy – or bad practice – method of factor analysis is? 
1. What are in short the differences between Principal Component Analyses and Exploratory Factor Analyses? 
1. What is the purpose of factor rotation? 


```{r echo=FALSE, results = 'asis', eval = config$show_text}
cat("#### Answers {-}

1. A latent variable is a variable (i.e. something that varies across our 
research subjects) that we cannot observe directly, but that we can 
infer by using some combination of observed variables that each tap a 
separate aspect of the latent variable (PL: own definition). Examples 
are IQ, personality, trust, math ability, etc. 
1. “Factor analysis (in the EFA context) is a method of discovering the 
number and nature of Latent variables that explain the variation and 
covariation in a set of measured variables” I..e. the method top 
investigate whether some set of observed variables together measure 
one or more latent variables.  
1. doing Principal component analysis, with no rotation, or varimax 
rotation, and retention of components with eigenvalues greater than 
1.0. These are the SPSS defaults, but they most often do not give a 
meaningful answer to a substantial research question. 
1. Main difference
    * PCA – summary of common variance (i.e. explained variance =100%) 
    * EFA – every factor has unique variance (i.e. explained variance =not 
100%) 
1. The procedure of rotating the factor axes to make sure items load as 
much on only one factor as possible There are two methods. 
Orthogonal rotation, in which two latent factors are not allowed to 
correlate (i.e. the axes describe a 90 degree angle), and oblique 
(oblimin or promax) rotation, in which the axes are allowed to 
correlate.  
")
```

#### Reference {-}

[Kestilä, Elina (2006) Is There Demand for Radical Right Populism in the 
Finnish Electorate? Scandinavian Political Studies 29(3),169-191](https://onlinelibrary.wiley.com/doi/abs/10.1111/j.1467-9477.2006.00148.x)

#### Questions {-}

1. What is the research question that the author tries to answer? 
2. Describe shortly the characteristics of the Radical Right Parties (RRP) in 
Europe. 
3. What are the two main explanations of support for RRP that this paper 
focuses on? Does the empirical part of the paper reflect the theoretical 
framework well?  
4. Is Finland very different from other European countries in on the main 
dependent variables according to the author? 
5. What is the conclusion by the author; i.e. what is the answer to the 
research question? 

```{r echo=FALSE, results = 'asis', eval = config$show_text}
cat("#### Answers {-}

1. There is no literal passage that can be used, but the research problem 
is stated as “why there is no Radical Right Party in Finland”. The 
question this paper poses is whether the attitudes the Finnish 
electorate holds about trust in politics and immigration is similar to or 
different from the other Western-European countries. 
2. pages 175-177 provide information on this. In short RRP emphasize 
anti-immigrant attitudes and connect immigrations to feelings of 
insecurity, spurring debates about crime. They are protest, anti-elitist 
parties, that often push for lower taxes, while protecting the welfare 
state. They can be clearly placed along the new cleavage in politics 
that is about materialism vs. post-materialism, and reflect the antonym 
of green parties. Finally, they represent the ‘losers’ of modernization 
and emphasize traditional moral values, social norms, the role of the 
family and the importance of a national culture. 
3. the anti-immigration issue and protest voting. 
4. No, Finland is very similar to other (Nordic) European countries. Some 
countries with the most anti-immigrant and anti-politics attitudes, still 
host a RRP. 
5. Anti-immigrant and anti-politics attitudes cannot explain why there is 
no RRP in Finland. The author find the reason for this in history, and 
the fact that politicians from mainstream political parties have caught 
the populist voters. 
")
```
